Sustainable Pathways for Irish homes in changing TIMES

### 2021-06-24 manuscript.tex
24-June-2021

\documentclass[preprint,12pt,authoryear]{elsarticle}
%\documentclass[10pt,final,journal,compsoc,a4paper]{IEEEtran}
%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

@@

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{pgf-pie}
\usepackage{calc}
\usepackage{graphicx}
\usepackage{float}
\usepackage{anyfontsize}
\usepackage{pgfplots}
\usepackage{amsmath}
\usepackage{comment}
% \usepackage{cite}

\usepackage[numeric-comp]{biblatex}

\addbibresource{references.bib}%Imports bibliography file

@@

\title{Sustainable Irish homes in changing TIMES}
%TIMES Ireland Model: Residential Sector
% The changing TIMES of Irish Homes
% Developing sustainable homes in changing TIMES

@@

@@

Ireland’s distinctive energy sectors are further separated from other EU member states when focusing on the residential sector. This is largely down to Ireland’s higher reliance on fossil fuels, lower building thermal efficiency, larger dwellings and rural settlement patterns. This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{EurostatShareSurveyilc_lvho50a}. The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts; (i) 2021 building regulations require a 70\% reduction of $CO_2$ emissions in new dwellings compared to dwellings built in 2005 \cite{DepartmentofHousingGov.ieBuildings}, and (ii) Central Statistics Office (CSO) data states approx. 85\% of residential buildings are 2005 or older. Furthermore, Ireland's fossil fuel dependency is underpinned by home heating oil accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.

@@

\subsection{Climate Targets}

@@

\subsection{Current Energy Policy}
\label{Intro:Policy}
Ireland has set out national policy objectives under 10 strategic investment priorities \cite{Ireland2018ProjectFramework}. The first strategic investment priority is ``Compact Growth" which aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, [as] a top priority"}. The densification of Ireland's main cities, will provide opportunities to improve the efficiency of heating, electricity and transport within and between cities. 
\par Specifically focusing on the energy sector, the Marginal Abatement Cost Curve (MACC) produced in CAP2019  \cite{EPA2018EPAReport} found electricity \& transport are most cost effective sectors for energy decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}. CAP2019 findings show the most cost effective solution to decarbonising the residential sector, is to receive higher amounts of low carbon electricity.

@@

Energy System Optimisation Models (ESOMs) can be used to explore least cost pathways, as they consider a large amount of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period. Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOM cover all the energy-related sectors, they can also be used to specifically explore the residential sector in higher detail, some examples are outlined in this section. 
\par For example, Li et al. \cite{Li2018IncorporatingModel} investigated the heating technology preference in UK, using UK TIMES Model (UKTM). As TIMES \cite{Loulou2016DocumentationI} assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}. Li set out to incorporate a Discrete Choice Model (DCM) into TIMES to better reflect how homeowners choose heating technologies. The DCM was developed from a nationwide survey with 442 homeowners completing the survey. The nationwide survey identified the most influential factors for determining homeowners' preferences for heating systems. The DCM then estimated the probability of a heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices. While Li's researcher has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights, through more surveys and analysis.

% Many bottom-up models have been used previously to specifically explore Ireland's residential sector \cite{Dineen2011ModellingHeating}\cite{Ahern2013StateStock}\cite{Ahern2015EnergySystems}\cite{Dineen2015ImprovedSource}\cite{Uidhir2020ResidentialIreland}\cite{Uidhir2020ImprovingDecisions} and some models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem}\cite{Connolly2021AalborgEnergy-System}, but very few models have used energy system optimisation model to explore the Irish residential sector. In 2014, Rogan et al used OSeMOYOS to explore energy service demands \cite{Rogan2014LEAPsSystem}, and Thellufsen examining DH with EnergyPLAN and TIMES in 2019 \cite{Thellufsen2019ImplementingIreland}.

@@

\subsection{Residential}
\label{LitR:RES}
The residential sector was developed in parallel with other sectors in TIM, and also been structured concurrently and similarly to the residential sector in the new Irish LEAP model, to allow more accessible soft-linking. Fig.\ref{fig:RES} shows the schematic overview of the residential sector in TIM, illustrating the flow of energy from fuel to demand.

\begin{figure}[!htbp]
 \centering
 \includegraphics[scale=0.45]{Figures/TIM_RES3.png}
 \caption{Residential Sector - Reference Energy System}
 \label{fig:RES}
\end{figure}

@@

The Archetype Dwelling Energy Model SQL (ArDEM-SQL) is a simulation model built on IS EN 13790, which converts the BER public database into Microsoft SQL language and then calculates annual energy consumption for space heating, water heating, pumps \& fans and lighting. ArDEM-SQL, is only used to change internal temperatures and that data is then used in TIM. NOTE large variation in energy rating due to occupancy behaviour - 
%Healy & Clinch reported that fuel-poor households were more likely to be experiencing colder temperatures than other households. In total almost 30% of fuelpoor households had living room temperatures below 18°C and more than two thirds had living room temperatures below 20°C. Overall nearly half of all non-fuel-poor homes had living room temperatures below 20°C. In households with persons aged over 65 years, 16% had living room temperatures below 18°C and over 50% had temperatures below 20°C. On the other end of the scale over 5% of households occupied by persons over 65 years experienced living room temperatures of ≥24°C.https://arrow.tudublin.ie/cgi/viewcontent.cgi?article=1089&context=scienmas
%The BER for the sample dwellings ranged from C1 to F. Over two thirds of the
sample dwellings had a BER of E or F. The remaining dwellings had a BER of C or
D. In total 4 of the sample dwellings had undergone energy efficiency upgrades in
recent years, including cavity wall insulation, attic insulation where applicable and
heating control upgrades. In the case of two sample dwellings, these upgrade
works were completed between monitoring period 1 and 2, and therefore results
could be compared before and after retrofit.

@@

@@

@@

\subsection{Strengths}
\label{Dis:Strengths}
\subsection{Limitations}
\label{Dis:Limit}
Consumers’ final investment decision can be
represented by “willingness-to-pay” curves – for
instance, around 10\% of owner-occupiers are willing
to invest in energy efficiency measures for a payback
time of 4 years, falling to 0\% for a payback time of
6 years for some consumer cohorts
\section{Conclusions}
\label{Conclusion}
This paper reports results on ambitious mitigation target in the period to 2050 for the Irish energy system. The analysis have been performed using the Irish TIMES model
\textit{We stress the importance of differentiating between technical con-
straints and behavioural constraints in optimisation models. Including both conflates what is considered desirable under the economic welfare paradigm (i.e. cost-minimising decisions) with “realistic decisions”, complicating the interpretation of the results. Optimisation models are by design well suited to study cost-optimal behaviour, and running scenarios just with technical constraints facilitates the identification of areas where market barriers are more entrenched, and interventions to accelerate the uptake of low-carbon practices may be needed.} \cite{Astudillo2017CanModelb}
\section{Acknowledgements}
\label{sec:Acknowledgements}
Tomas %CAPACITY is funded under the Climate Action Modelling Group (CAMG) by DECC
% https://www.marei.ie/project/capacity/
CHIMERA is supported by a research grant from Science Foundation Ireland (SFI) and the National Natural Science Foundation of China (NSFC) under the SFI-NSFC Partnership Programme, grant no. 17/NSFC/5181 and
supported by MaREI, the SFI Research Centre for Energy, Climate, and Marine [Grant No: 12/RC/2302\_P2]

>>>>

\documentclass[authoryear,preprint,12pt]{elsarticle}
%\documentclass[10pt,final,journal,compsoc,a4paper]{IEEEtran}
%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

@@

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{pgf-pie}
\usepackage{calc}
\usepackage{graphicx}
\usepackage{float}
\usepackage{anyfontsize}
\usepackage{pgfplots}
\usepackage{amsmath}
\usepackage{comment}
%\usepackage{cite}

\usepackage{biblatex}
\addbibresource{references.bib} %Imports bibliography file

@@

\title{Sustainable Pathways for Irish homes in changing TIMES}
%TIMES Ireland Model: Residential Sector
% The changing TIMES of Irish Homes
% Developing sustainable homes in changing TIMES

@@

Energy System Optimisation Models (ESOMs) can be used to explore least cost pathways, as they consider a large amount of variables to satisfy demand, while accounting for complex sectoral interconnections within each time period. Some of the main ESOM frameworks include EnergyPLAN \cite{stergaard2015ReviewingSimulations}, MESSAGE \cite{Messner1995WorkingE}, TIMES \cite{Loulou2016DocumentationI}, Balmorel \cite{Wiese2018BalmorelModel}, TEMOA \cite{Hunter2013ModelingTemoa} and OSeMOSYS \cite{Howells2011OSeMOSYS:Development.}.While ESOM cover all the energy-related sectors, they can also be used to specifically explore the residential sector in higher granularity, some international examples are outlined in this section. 
\par For example, Li et al. \cite{Li2018IncorporatingModel} investigated the heating technology preference in UK, using UK TIMES Model (UKTM). As TIMES \cite{Loulou2016DocumentationI} assumes perfect foresight and \textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel}. Li set out to incorporate a Discrete Choice Model (DCM) into TIMES to better reflect how homeowners choose heating technologies. The DCM was developed from a nationwide survey with 442 homeowners completing the survey. The nationwide survey identified the most influential factors for determining homeowners' preferences for heating systems. The DCM then estimated the probability of a heating technology selection and cost was found not to have a statistically significant impact on homeowners' choices. While Li's researcher has provided some initial insights, future work can improve the robustness and comprehensiveness of the insights, through more surveys and analysis.

%The Cambridge Housing Model (BEIS) ses English Housing Survey data, coupled to a SAP ( UK's BER equivalent ) energy calculator, to estimate energy use and CO2 emissions for all homes in England, broken down by final use. The CHM reads in the EHS dwelling for each case and performs building physics calculations to determine energy consumption and associated CO2 emissions, by use and by fuel type. Multiplying the energy use and CO2 emissions by the associated weighting and summing across all cases gives total values for England. Housing Data is cleaned, Climate data for region is 8 energy services - space heating ( main \& secondary ), water heating, space cooling, lighting, electrical appliances, cooking and pumps \& fan. 6 fuels - gas, oil, solid, biomass, electriicty and renewable

\subsubsection{Ireland: Residential Modelling}
\label{Intro:ESOMIreland}

In Ireland, previous modelling on the residential sector has often involved bottom-up residential models, rather than energy system models specifically examining the residential sector. Nevertheless, the residential models have provided insights on input data and methodology approach used in Ireland. In 2011 Dineen and Ó Gallachóir \cite{Dineen2011ModellingHeating} developed a residential bottom-up model to assess the impacts of measures in Ireland's National Energy Efficiency Action Plan (NEEAP). The model focused on energy efficiency of space and water heating by examining technology and building fabric efficiency. The building fabric efficiency of newly occupied dwellings developed takes into account the theoretical increase in energy efficiency due to improved 2008 and 2010 building regulations. Space heating demand accounted for 70\% of household energy consumption in newly built dwellings in 1997, down to just under 60\% in 2006 and the model is predicts a reduction to 27\% by 2015.The paper acknowledges retrofitting of existing stock will play a large role in meeting energy efficiency targets, however it is not examined. Nor is the affect of internal temperatures on energy efficiency improvements.  
\par In 2013 Ahern et al.\cite{Ahern2013StateStock} investigated thermal retrofit measures to the existing detached, oil centrally heated, rural housing stock. A base geometry according to age bands and thermal characteristics was created in the study. The heat load for a statistical occupancy and year was established. The default values used in some of the BER database variables where highlighted. For example,  there are assumed U-values when the roof insulation thickness is unknown, this is based on age, but there are no U-values for uninsulated roofs and a 2001 survey found 18\% of detached housing in Ireland had no roof insulation. The internal temperature assumed for all dwelling was 20$^{\circ}$C. Ahern et al. found that newer dwellings, consume the greatest amount of energy in kWh/annum than all the older dwellings, due to the fact that their large floor area. Concluding the majority of energy efficiency progress for thermal use in Irish dwellings is being offset.

The reduction in heating costs and CO2 emissions for houses constructed prior to 1979 (coinciding with the introduction of building regulations) and around 26% for newer homes, thus offering a significant contribution (44%) to Ireland’s residential carbon abatement projections and hence in meeting the EU’s directives on energy and carbon.

through energy efficiency measures has been offset by the so called size effect. The
The model predicts a reduction of 65% in running costs and
CO2 emissions for house types A–E and a reduction of 26% in running costs and CO2 emissions for house types F–J from retrofit measures, and forecasts a total saving of 276 million euro and a CO2 abatement of 1.21 MtCO2. By
the
potential exists to reduce running costs and CO2 emissions by an average of 63% for housing constructed prior to the 1979 building regulations (house types A–E) and by 26% for newer housing. The ratio of amounts of CO2 which can be potentially saved to
the amount of money saved was calculated as 4.35 kgCO2/h. Nationally an annual saving of h183 million euro is theoretically achievable with a resultant saving of 0.80 MtCO2 arising from insulation retrofits, the replacement of single glazed windows and increasing air tightness. The
\cite{Ahern2015EnergySystems}\cite{Dineen2015ImprovedSource}\cite{Uidhir2020ResidentialIreland}\cite{Uidhir2020ImprovingDecisions} and some models have been used to optimise Ireland's entire energy system \cite{Chiodi2013ModellingSystem}\cite{Connolly2021AalborgEnergy-System}, but very few models have used energy system optimisation model to explore the Irish residential sector. In 2014, Rogan et al used OSeMOYOS to explore energy service demands \cite{Rogan2014LEAPsSystem}, and Thellufsen examining DH with EnergyPLAN and TIMES in 2019 \cite{Thellufsen2019ImplementingIreland}.

@@

Ireland’s distinctive energy sectors are further separated from other EU member states when focusing on the residential sector. This is largely down to Ireland’s higher reliance on fossil fuels, lower building thermal efficiency, larger dwellings and rural settlement patterns. This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{EurostatShareSurveyilc_lvho50a}. The low thermal efficiency of Ireland's dwelling stock is highlighted by two facts; (i) 2021 building regulations require a 70\% reduction of CO\textsubscript{2} emissions in new dwellings compared to dwellings built in 2005 \cite{DepartmentofHousingGov.ieBuildings}, and (ii) Central Statistics Office (CSO) data states approx. 85\% of residential buildings are 2005 or older. Furthermore, Ireland's fossil fuel dependency is underpinned by home heating oil accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.

@@

\subsubsection{Climate Targets}

@@

\subsubsection{Current Energy Policy}
\label{Intro:Policy}
Ireland has set out national policy objectives under 10 strategic investment priorities \cite{Ireland2018ProjectFramework}. The first strategic investment priority is ``Compact Growth" which aims to  achieve \textit{``effective density and consolidation, rather than more sprawl of urban development, [as] a top priority"}. The densification of Ireland's main cities, will provide opportunities to improve the efficiency of heating, electricity and transport within and between cities. 
\par Specifically focusing on the energy sector, the Marginal Abatement Cost Curve (MACC) produced in CAP2019  \cite{EPA2018EPAReport} found electricity \& transport are most cost effective sectors for energy decarbonisation \cite{DepartmentofCommunicationsClimateActionandEnvironment2019DecarbonisationIreland}. CAP2019 findings show the most cost effective solution to decarbonising the residential sector, is to receive higher amounts of low carbon electricity.

@@

@@

\subsection{Residential}
\label{LitR:RES}
The residential sector was developed in parallel with other sectors in TIM, and also been structured concurrently and similarly to the residential sector in the new Irish LEAP model, to allow more accessible soft-linking. Fig.\ref{fig:RES} shows the schematic overview of the residential sector in TIM, illustrating the flow of energy from fuel to demand.  
First iteration - of model will need further development DH, BEhavour, higher granularity and spatial will be easily done with dataset. 
\begin{figure}[!htbp]
 \centering
 \includegraphics[scale=0.45]{Figures/TIM_RES3.png}
 \caption{Residential Sector - Reference Energy System}
 \label{fig:RES}
\end{figure}

@@

The Archetype Dwelling Energy Model SQL (ArDEM-SQL) is a simulation model built on IS EN 13790, it converts the BER public database into Microsoft SQL language and then calculates annual energy consumption for space heating, water heating, pumps \& fans and lighting. ArDEM-SQL is used to clean the input data in TIM, by changing internal temperatures. The internal temperature is highly dependent on occupancy behaviour, a thermal comfort  thermal discomfort are higher among the over-65s than for other household \cite{Healy2002FuelIreland} - living room temperatues between 12 and 26...s  and therefore there is are large variations per BER rating.
The International Standard ISO 7730 [6] uses these PMV and PPD indices to predict the thermal sensation of people exposed to moderate thermal environments, as well as to specify acceptable thermal environmental conditions for comfort...
Despite this, it is
worth noting that the most inefficient dwellings i.e. band F, managed to achieve an
average temperature during period 2 that was only 0.5°C lower than the most
efficient dwellings i.e. band C.
C = 19.2, D = 18, E = 18.2 and F= 18.7 older occupants dublin Ho
%Healy & Clinch reported that fuel-poor households were more likely to be experiencing colder temperatures than other households. In total almost 30% of fuelpoor households had living room temperatures below 18°C and more than two thirds had living room temperatures below 20°C. Overall nearly half of all non-fuel-poor homes had living room temperatures below 20°C. In households with persons aged over 65 years, 16% had living room temperatures below 18°C and over 50% had temperatures below 20°C. On the other end of the scale over 5% of households occupied by persons over 65 years experienced living room temperatures of ≥24°C.https://arrow.tudublin.ie/cgi/viewcontent.cgi?article=1089&context=scienmas
%The BER for the sample dwellings ranged from C1 to F. Over two thirds of the
sample dwellings had a BER of E or F. The remaining dwellings had a BER of C or
D. In total 4 of the sample dwellings had undergone energy efficiency upgrades in
recent years, including cavity wall insulation, attic insulation where applicable and
heating control upgrades. In the case of two sample dwellings, these upgrade
works were completed between monitoring period 1 and 2, and therefore results
could be compared before and after retrofit.

@@

EPC register EU - Ireland
%The quality of Swedish EPC data was initially studied by Göransson [41], subsequently by Claesson [42], Mangold et al. [21] and Hårsman et al. [43], and most recently by Pasichnyi et al. [6]. These studies have found a couple of important deficiencies in Swedish EPC data. First, the energy expert, and the firm he or she belongs to, has been shown to generate variations in Swedish EPC data [43], [44]. More specifically, Hårsman et al. [43] found that the energy expert along with the characteristics of the firm caused a variation of ±20% in the assessment of energy consumption.
%Second, through interviews with energy experts, Claesson [42] found that most EPCs contain uncertain estimates of energy use as well as arbitrarily distributed values—thus challenging the notion that Swedish EPC data quality, while allegedly measured, should be superior to the quality of calculated values in other member states.

@@

no elasticity - fuel prices and weather compensation?

@@

\subsection{Strengths}
\label{Dis:Strengths}

Detailed bottom up model - using 2018 base year ( in line with policy )
Updated technology
Combines energy system effects
Robust filtered input data about housing stock

\subsection{Limitations}
\label{Dis:Limit}
Consumers’ final investment decision can be
represented by “willingness-to-pay” curves – for
instance, around 10\% of owner-occupiers are willing
to invest in energy efficiency measures for a payback
time of 4 years, falling to 0\% for a payback time of
6 years for some consumer cohorts
Type of Tenure is ignored
BER aggeragateed
housing stock u-values
behaviour and occupancy
Area
\section{Conclusions}
\label{Conclusion}
This paper reports results on ambitious mitigation target in the period to 2050 for the Irish energy system. The analysis have been performed using the Irish TIMES model
\textit{We stress the importance of differentiating between technical con-
straints and behavioural constraints in optimisation models. Including both conflates what is considered desirable under the economic welfare paradigm (i.e. cost-minimising decisions) with “realistic decisions”, complicating the interpretation of the results. Optimisation models are by design well suited to study cost-optimal behaviour, and running scenarios just with technical constraints facilitates the identification of areas where market barriers are more entrenched, and interventions to accelerate the uptake of low-carbon practices may be needed.} \cite{Astudillo2017CanModelb}
\section{Acknowledgements}
\label{sec:Acknowledgements}
Tomas %CAPACITY is funded under the Climate Action Modelling Group (CAMG) by DECC
% https://www.marei.ie/project/capacity/
CHIMERA is supported by a research grant from Science Foundation Ireland (SFI) and the National Natural Science Foundation of China (NSFC) under the SFI-NSFC Partnership Programme, grant no. 17/NSFC/5181 and
supported by MaREI, the SFI Research Centre for Energy, Climate, and Marine [Grant No: 12/RC/2302\_P2]
